The Talking Heads Experiment

### 2015-04-27 chap12/chap12.tex
changes to headings and minor mistakes across several chapters
hyphenation, punctuation, positioning, etc for chapters 1-6
fixed remaining typos

\chapter{Language Evolution}
\label{c:evolution}

@@

\centerline{
\includegraphics[height=0.5\textwidth]{chap12/figs/full-1000.pdf}$~~~$
\includegraphics[height=0.5\textwidth]{chap12/figs/full-3000.pdf}
}
\caption{{The {\itshape Basic Colour Strategy} allows a population of agents (in this case 10) to self-organise a colour lexicon from scratch. The graph shows that high communicative success is reached with a lexicon of about 15 colour words (top). The evolution of a typical lexicon in a smaller population (5 agents), is shown after 400 (bottom left) and 1200 (bottom right) games per agent. Each row represents the lexicon of one agent.}
\label{fig:full-cs-dynamics}
} 
\end{figure}

@@

\centerline{
\includegraphics[height=0.5\textwidth]{chap12/figs/bw-1000.pdf} $~~~$
\includegraphics[height=0.5\textwidth]{chap12/figs/bw-4000.pdf}}
\label{fig:brightness-dynamics}
\caption{{The {\itshape Brightness Colour Strategy} also allows a population of agents (in this case 10)
to evolve an adequate colour lexicon (top). The evolution of a typical lexicon in a smaller population (5 agents), is shown 
after 400 (bottom left) and 1600 (bottom right) games per agent. Each row represents
the lexicon of one agent.}}
\end{figure}

@@

We will use the domain of spatial language again as a source of examples, based on experiments carried out by 
Michael Spranger. It uses the \textsc{qrio} humanoid robots 
and scenes like the one shown in \figref{fig:spatial}. An example of a constraint network from this domain 
is shown in \figref{fig:spatial-strategies}. It is the network for the phrase `near the box'. `Near' is a proximal 
relation which categorises objects in terms of how near they are to a landmark (in this case the box). In a list form the 
network looks like this: 
\begin{small}
\begin{verbatim}
1. (get-context ?ctx-1) 
2. (construct-region-proximal 
    ?region ?ctx-1 ?landmark ?cat) 
3. (bind proximal-category ?cat NEAR) 
4. (apply-selection ?landmark ?landmarks ?selector-2)
5. (apply-class ?landmarks ?ctx ?landmark-class) 
6. (bind selector ?selector-2 UNIQUE) 
7. (get-context ?ctx) 
8. (bind object-class ?landmark-class BOX)
\end{verbatim}
\end{small}
The order of these operations does not matter as values propagate and constraints become active until no further progress
can be made. Operations in 4, 5, 6, 7 and 8 provide the referent of `the box'. The following contributions 
can be made by each of the cognitive operations: The predicate \textsc{box} gets bound to ?landmark-class in operation 8. Object-class is 
the type of ?landmark-class. The context, i.e. all objects in the present situation, gets bound to ?ctx in operation 7. 
The selector \textsc{unique} gets bound to ?selector-2 in operation 6. 
Then the operation apply-class is applied to all the objects in ?ctx (the context) to find the possible landmarks which 
get bound to ?landmarks using the ?landmark-class (\textsc{box}). This set is assumed to be a singleton and so 
in operation 4 the unique element in the set of ?landmarks can be selected 
and bound to ?landmark. 
\enlargethispage{1\baselineskip}
The remaining three operations apply the proximal relation. In operation 3 ?cat is bound to the proximal-category \textsc{near}. 
In operation 1, ?ctx-1 gets bound to 
the context. And finally in operation 2, the region around the landmark, which satisfies the category ?cat (\textsc{near}), can 
be computed and bound to ?region by ``construct-region-proximal''. The construction-region-proximal constraint has to perform first 
a geometric transformation on the scene, constructing a world model from the viewpoint of the landmark, i.e. the box.

A {\itshape conceptualisation strategy} is an abstraction of such a concrete conceptualisation network. 
\is{conceptualisation strategy}
It consists of a network that acts as a useful chunk and is therefore a modular unit that can be linked into a larger network. 
The chunk has a set of open variables with which the network can be linked to other networks. 
For example, the network shown in \figref{fig:spatial-strategies} can be made more abstract to handle many instances 
of proximal relations, not only `near the box', but also `far from the box', `close to the box', `near the boxes', 
`near the table', `very far from the wall', etc. 
\enlargethispage{1em}
The open variables are ?cat, ?selector-2, and ?landmark-class, which have to be supplied by lexical items or 
phrases. The network is then as follows, with operations at 3, 6 and 8 removed: 
\begin{small}
\begin{verbatim}
1. (get-context ?ctx-1) 
2. (construct-region-proximal ?region ?ctx-1 ?landmark ?cat) 
4. (apply-selection ?landmark ?landmarks ?selector-2)
5. (apply-class ?landmarks ?ctx ?landmark-class) 
7. (get-context ?ctx) 
\end{verbatim}
\end{small}
The network has to be completed with bindings for the open variables, but all the other variables can be 
computed through the network itself. 
Network-chunks are expressed through a grammatical construction, which expresses which chunk is involved 
and how the initial values of the slots are to be expressed. In this 
case it is a noun-phrase construction, syntactically of the form [noun preposition article noun] and 
semantically constrained as [proximal-category selector object-class].

@@

Historical linguists have researched many concrete examples of how agreement systems arise.\footnote{See for example: \cite{Lehmann:1988}, \cite{Luraghi:2011}.}
There are invariably the following phases: 
\begin{enumerate}
\item Agreement markers derive from reusing existing words, such as pronouns or classifiers. 
\item The markers derived from independent words become shorter, they lose part of their form, 
then become clitics and later affixes. For example, the Diyrbal (Australian aboriginal)
agreement marker {\itshape m-} evolved from the classifier {\itshape mayi}, which means 'non-flesh food'. 
\item The features that agreement markers express, which are initially semantically 
grounded, invariably become more abstract and get to be used in 
a purely conventional manner. For example, the masculine/feminine 
gender distinction has its basis in male/female sex, but is then arbitrarily applied to inanimate objects, so that table 
might be masculine in one language (German: {\itshape der Tisch}) but feminine in another (French: {\itshape la table}). 
\end{enumerate}
Starting from a game of reference, similar to the Talking Heads experiment, populations of agents were programmed 
first to build a ``formal'' agreement system, i.e. a system based on formal markers without any meaning, and then 
a meaningful grammatical system, with grammatical markers that express semantic features. The experiment was 
then extended with further strategies that optimise three niche factors:
\begin{enumerate}
\item The recruitment of existing words as markers makes the acquisition of new markers more accurate because it is based 
on existing words and so the meaning can be guessed more easily. This helps to propagate the markers faster in the population. 
\item The phonological reduction of wordforms leads to less articulatory effort, as we have already seen with the German 
article system. 
\item Conventionalisation leads to a smaller inventory, and hence less memory requirements and better preservation of the 
language system. 
\end{enumerate}

@@

First we need much more work along the paths of which we have seen many examples: 
\begin{enumerate}
\item The study of more language systems, based on firm empirical data. For example, the study of the verbal system of 
Italian or the determiner system of Dutch. 
\item A study of the language strategies that underly these systems. A language strategy should define how a particular 
type of language system operates, how it is learned, and how it is expanded and aligned. 
\item A study of the niche occupied by the language systems governed by a particular language strategy. 
\item Evolutionary experiments to show how new strategies may arise. 
\end{enumerate} 
Then there are also several deep open questions which have not been addressed extensively so far. Here are a few examples: 
\begin{enumerate}
\item What are implications of the models presented here for the neurobiology of language? Right now it remains largely 
a mystery how the human brain is processing language and current neural network models are entirely insufficient from 
a representational and computational point of view to handle the structures and processes that we have been using in 
the various experiments reported in this book. 
\item How to orchestrate long-term evolution, both within the individual and in a population? Some suggestions have been 
made based on the autotelic principle which allows agents to control the communicative challenges in 
relation to their level of skill \citep{Steels:2007}. 
But these suggestions need to be tested
more seriously in large-scale long term experiments. 
\item What is the interaction between cultural evolution and genetic evolution? I have argued in favour of culture-driven genetic
evolution but what kind of genetic change is necessary to support the processes that we have been found to 
be operating at the cultural level? Clearly, flexible recruitment and huge amounts of memory are key demands on the 
human brain, but there are undoubtly others. 
\item What is the interaction between social evolution, cultural evolution and biological evolution? The language 
games which we have studied all required honest communication. Agents do their best to play the game as well as they 
can and agents are fully cooperative. This is not the default condition in the natural world and it is not so easy 
to explain how human sociality arose, and how human language may have fostered it \citep{Dor:2014}. 
\end{enumerate}
These questions can only be properly addressed by bringing up and working out hypotheses and
doing detailed computational analysis and computational 
experiments. The field of evolutionary linguistics is still in its infancy and so far too few researchers
are contributing through the kind of computational simulations that were discussed in this book. This is not surprising given 
the technical complexity and the breadth of knowledge needed to work in this area. It takes Ph.D students on average 5
years of training and experimentation to carry a research thread to 
fruition, assuming that the student starts with considerable background in computational linguistics or Artificial Intelligence.
Let us hope that in the future the resources will be available so that younger generations can pursue this exciting path. 
Many discoveries are waiting to be made!

>>>>

\chapter{Language evolution}
\label{c:evolution}

@@

\centerline{
\includegraphics[height=0.5\textwidth]{chap12/figs/full-1000.pdf} ~~~
\includegraphics[height=0.5\textwidth]{chap12/figs/full-3000.pdf}
}
\caption{{The {\itshape Basic Colour Strategy} allows a population of agents (in this case 10) to self-organise a colour lexicon from scratch. The graph shows that high communicative success is reached with a lexicon of about 15 colour words (top). The evolution of a typical lexicon in a smaller population (5 agents), is shown after 400 (bottom left) and 1200 (bottom right) games per agent. Each row represents the lexicon of one agent.}
\label{fig:full-cs-dynamics}
} 
\end{figure}

@@

\centerline{
\includegraphics[height=0.5\textwidth]{chap12/figs/bw-1000.pdf} ~~~
\includegraphics[height=0.5\textwidth]{chap12/figs/bw-4000.pdf}}
\label{fig:brightness-dynamics}
\caption{{The {\itshape Brightness Colour Strategy} also allows a population of agents (in this case 10)
to evolve an adequate colour lexicon (top). The evolution of a typical lexicon in a smaller population (5 agents), is shown 
after 400 (bottom left) and 1600 (bottom right) games per agent. Each row represents
the lexicon of one agent.}}
\end{figure}

@@

We will use the domain of spatial language again as a source of examples, based on experiments carried out by 
Michael Spranger. It uses the \textsc{qrio} humanoid robots 
and scenes like the one shown in \figref{fig:spatial}. An example of a constraint network from this domain 
is shown in \figref{fig:spatial-strategies}. It is the network for the phrase `near the box'. `Near' is a proximal 
relation which categorises objects in terms of how near they are to a landmark (in this case the box). In a list form the 
network looks like this: 
\begin{small}
\begin{verbatim}
1. (get-context ?ctx-1) 
2. (construct-region-proximal 
    ?region ?ctx-1 ?landmark ?cat) 
3. (bind proximal-category ?cat NEAR) 
4. (apply-selection ?landmark ?landmarks ?selector-2)
5. (apply-class ?landmarks ?ctx ?landmark-class) 
6. (bind selector ?selector-2 UNIQUE) 
7. (get-context ?ctx) 
8. (bind object-class ?landmark-class BOX)
\end{verbatim}
\end{small}
The order of these operations does not matter as values propagate and constraints become active until no further progress
can be made. Operations in 4, 5, 6, 7 and 8 provide the referent of `the box'. The following contributions 
can be made by each of the cognitive operations: The predicate \textsc{box} gets bound to ?landmark-class in operation 8. Object-class is 
the type of ?landmark-class. The context, i.e. all objects in the present situation, gets bound to ?ctx in operation 7. 
The selector \textsc{unique} gets bound to ?selector-2 in operation 6. 
Then the operation apply-class is applied to all the objects in ?ctx (the context) to find the possible landmarks which 
get bound to ?landmarks using the ?landmark-class (\textsc{box}). This set is assumed to be a singleton and so 
in operation 4 the unique element in the set of ?landmarks can be selected 
and bound to ?landmark.

\enlargethispage{1\baselineskip}
The remaining three operations apply the proximal relation. In operation 3 ?cat is bound to the proximal-category \textsc{near}. 
In operation 1, ?ctx-1 gets bound to 
the context. And finally in operation 2, the region around the landmark, which satisfies the category ?cat (\textsc{near}), can 
be computed and bound to ?region by ``construct-region-proximal''. The construction-region-proximal constraint has to perform first 
a geometric transformation on the scene, constructing a world model from the viewpoint of the landmark, i.e. the box.

A {\itshape conceptualisation strategy} is an abstraction of such a concrete conceptualisation network. 
\is{conceptualisation strategy}
It consists of a network that acts as a useful chunk and is therefore a modular unit that can be linked into a larger network. 
The chunk has a set of open variables with which the network can be linked to other networks. 
For example, the network shown in \figref{fig:spatial-strategies} can be made more abstract to handle many instances 
of proximal relations, not only `near the box', but also `far from the box', `close to the box', `near the boxes', 
`near the table', `very far from the wall', etc. 
The open variables are ?cat, ?selector-2, and ?landmark-class, which have to be supplied by lexical items or 
phrases. The network is then as follows, with operations at 3, 6 and 8 removed: 
\begin{small}
\begin{verbatim}
1. (get-context ?ctx-1) 
2. (construct-region-proximal ?region ?ctx-1 ?landmark ?cat) 
4. (apply-selection ?landmark ?landmarks ?selector-2)
5. (apply-class ?landmarks ?ctx ?landmark-class) 
7. (get-context ?ctx) 
\end{verbatim}
\end{small}
The network has to be completed with bindings for the open variables, but all the other variables can be 
computed through the network itself. 
Network-chunks are expressed through a grammatical construction, which expresses which chunk is involved 
and how the initial values of the slots are to be expressed. In this 
case it is a noun-phrase construction, syntactically of the form [noun preposition article noun] and 
semantically constrained as [proximal-category selector object-class].

@@

Historical linguists have researched many concrete examples of how agreement systems arise.\footnote{See for example: \cite{Lehmann:1988}, \cite{Luraghi:2011}.}
There are invariably the following phases: 
\begin{enumerate}
\item Agreement markers derive from reusing existing words, such as pronouns or classifiers. 
\item The markers derived from independent words become shorter, they lose part of their form, 
then become clitics and later affixes. For example, the Dyirbal (Australian aboriginal)
agreement marker {\itshape m-} evolved from the classifier {\itshape mayi}, which means 'non-flesh food'. 
\item The features that agreement markers express, which are initially semantically 
grounded, invariably become more abstract and get to be used in 
a purely conventional manner. For example, the masculine/feminine 
gender distinction has its basis in male/female sex, but is then arbitrarily applied to inanimate objects, so that table 
might be masculine in one language (German: {\itshape der Tisch}) but feminine in another (French: {\itshape la table}). 
\end{enumerate}
Starting from a game of reference, similar to the Talking Heads experiment, populations of agents were programmed 
first to build a ``formal'' agreement system, i.e. a system based on formal markers without any meaning, and then 
a meaningful grammatical system, with grammatical markers that express semantic features. The experiment was 
then extended with further strategies that optimise three niche factors:
\begin{enumerate}
\item The recruitment of existing words as markers makes the acquisition of new markers more accurate because it is based 
on existing words and so the meaning can be guessed more easily. This helps to propagate the markers faster in the population. 
\item The phonological reduction of wordforms leads to less articulatory effort, as we have already seen with the German 
article system. 
\item Conventionalisation leads to a smaller inventory, and hence less memory requirements and better preservation of the 
language system. 
\end{enumerate}

@@

First we need much more work along the paths of which we have seen many examples: 
\begin{enumerate}
\item The study of more language systems, based on firm empirical data. For example, the study of the verbal system of 
Italian or the determiner system of Dutch. 
\item A study of the language strategies that underly these systems. A language strategy should define how a particular 
type of language system operates, how it is learned, and how it is expanded and aligned. 
\item A study of the niche occupied by the language systems governed by a particular language strategy. 
\item Evolutionary experiments to show how new strategies may arise. 
\end{enumerate} 
Then there are also several deep open questions which have not been addressed extensively so far. Here are a few examples: 
\begin{enumerate}
\item What are implications of the models presented here for the neurobiology of language? Right now it remains largely 
a mystery how the human brain is processing language and current neural network models are entirely insufficient from 
a representational and computational point of view to handle the structures and processes that we have been using in 
the various experiments reported in this book. 
\item How to orchestrate long-term evolution, both within the individual and in a population? Some suggestions have been 
made based on the autotelic principle which allows agents to control the communicative challenges in 
relation to their level of skill \citep{Steels:2007}. 
But these suggestions need to be tested
more seriously in large-scale long term experiments. 
\item What is the interaction between cultural evolution and genetic evolution? I have argued in favour of culture-driven genetic
evolution but what kind of genetic change is necessary to support the processes that we have been found to 
be operating at the cultural level? Clearly, flexible recruitment and huge amounts of memory are key demands on the 
human brain, but there are undoubtedly others. 
\item What is the interaction between social evolution, cultural evolution and biological evolution? The language 
games which we have studied all required honest communication. Agents do their best to play the game as well as they 
can and agents are fully cooperative. This is not the default condition in the natural world and it is not so easy 
to explain how human sociality arose, and how human language may have fostered it \citep{Dor:2014}. 
\end{enumerate}
These questions can only be properly addressed by bringing up and working out hypotheses and
doing detailed computational analysis and computational 
experiments. The field of evolutionary linguistics is still in its infancy and so far too few researchers
are contributing through the kind of computational simulations that were discussed in this book. This is not surprising given 
the technical complexity and the breadth of knowledge needed to work in this area. It takes Ph.D students on average 5
years of training and experimentation to carry a research thread to 
fruition, assuming that the student starts with considerable background in computational linguistics or Artificial Intelligence.
Let us hope that in the future the resources will be available so that younger generations can pursue this exciting path. 
Many discoveries are waiting to be made!

### 2015-04-27 chap2/chap2.tex
fixed remaining typos

Clearly in the development of language and meaning, 
the group and the environment matter. A child 
who grows up without a caring 
community or without sufficient environmental stimuli
never develops the rich cognitive capacities 
normal adults have. From attempts to educate {\it wolf children}
who grow up in isolation from a human community, or 
impaired children for whom the intensity of early 
interactions are limited, we know that 
there are critical periods
where a community and a challenging environment must be
present otherwise the child's capacities for language
are damaged for the 
rest of his or her life.\footnote{See \cite{Tager:1994}.}

But how can we sufficiently 
recreate these social conditions in experiments 
with artificial systems? 
Building colonies of physical autonomous robots
roaming the world in search of stimulating 
environments and rich interactions with
other robots is not feasable today. So how 
can we ever test seriously situated and 
socially embedded approaches to cognition?

@@

Let me make a distinction between the physical aspects of 
a cognitive agent and the mental aspects. The physical aspects include the 
agent's body, the sensors and articulators, 
the physical location, the objects in this location, and the other
agents physically present in the environment. The 
mental aspects include the agent's repertoire of behaviours, 
the brain structures and processes
performing categorisations of reality, his
memory, lexicon, grammar, and so on. 
In the case of humans, these two aspects are intimately 
connected and indivisible. We cannot teleport our mental 
faculties into another body, or into another copy of 
our body, even though there have been speculations that 
we could in the future record human 
brain states,\footnote{
Such visions of the future have been put forward
by \cite{Moravec:1995}. Neither current artificial intelligence technology nor
the state of the art in brain state recording are anywhere 
near to realising these visions.}
I believe that this will still not enable teleportation
because in human brains there is no distinction between 
hardware and software. The architecture of a human brain, 
the physical connections between cells, and the 
biochemical processes in each cell determine the brain's behaviour.
There is no separation between a {\it brain program} and
an interpreter that reads brain programs and 
executes them. The brain is a special-purpose 
hardware device which is unique to each individual. 
To copy such a device we would have to rebuild it physically,
atom by atom, and integrate it in an exact copy of the 
same body.

@@

A blind person who receives sight after the critical 
period for acquiring visual categorisation undergoes a 
traumatic experience, see \cite{Zeki:1993}. 
We could in principle build robot bodies
through which agents can experience their world with tactile
sensing and other robot bodies which support visual sensing, or both. 
But then an agent which had only access to tactile
sensing might suddenly find himself in a body equiped with 
vision. This pathological complication will 
be avoided by using the same robotic infrastructure
in every location, even though it would be fascinating and 
technically possible to study multi-modality in its 
own right. 
We also decided to make the robots vision-based, because visual 
sensing is one of the major sources of meaning in
human natural languages.

@@

The Guessing Game is one of the common things we do with language. 
For example, I play a similar game when I sit with a friend
at the dining table and say `could you give me the salt'. 
If she guesses correctly what I mean and hands 
me the salt, the game has succeeded. If she looks at me 
with a puzzled face (maybe she does not speak English) or hands me
the salmon instead of the salt, the game has failed. In that case, 
I can gesture in the direction of the salt and say `no, no, the {\itshape salt}
please', and then she hopefully realises and gives the salt to me. Failure 
is common in natural language dialogues and may be caused by 
many factors. For example, the salmon could indeed 
be close to the salt and my pronounciation of the word `salt'
may have sounded a bit like `salmon', perhaps because there was
loud music playing in the background or perhaps
my friend does not understand English. 
A failure is often 
an opportunity to negotiate how something will be expressed
in the future. For example, the hearer may pick up a new word
or the speaker may realise that a certain word is 
not appropriate in this particular context.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap2/figs/triangle.pdf}}
\caption{\label{triangle}Any verbal interaction
involves four entities here grouped in the
{\it semiotic square}. The relation between utterance
and referent always needs to be established indirectly by 
passing through perception and meaning.}
\end{figure}

@@

Every physical Talking Heads set-up features a {\it waiting room} in 
which agents are stored, ready to be loaded into the 
robotic infrastructure or to be teleported to another
physical site. A game starts when two agents are chosen
at random from this waiting room and loaded into 
the two physical robots. The
internal architecture of each agent is connected with 
the sensori-motor apparatus of the robot so that they can 
receive the sensory data streams recorded by the camera
and control the movements of the robots. Then the general 
control system assigns randomly the role of speaker and hearer
to each of the agents and the game can begin.

@@

The data on sensory channels are values from a continuous domain
(between 0.0 and 1.0). 
To be the basis of natural language communication, these
values must be transformed into a discrete domain. This is 
precisely the task of the conceptual layer. It 
performs massive data reduction to make infinitely
rich environments managable. One means of categorisation
is to divide up each domain of values on a particular
sensory channel into regions and 
simply assign a category to each region, thus creating a 
discrimination tree.\is{discrimination tree} For example, 
the \textsc{hpos}-channel can be cut in two halves leading to a 
distinction between [\textsc{left}] ($0.0 ≤q \textsc{hpos} < 0.5$) and [\textsc{right}] 
($0.5 ≤q \textsc{hpos} ≤q 1.0$). The triangular object 
in \figref{scene1-1} has the value \textsc{hpos}=0.35 
and would therefore be categorised as [\textsc{left}]. Similarly, the 
\textsc{vpos}-channel can be divided in two halves yielding the 
categories [\textsc{lower}] and [\textsc{upper}], and likewise the \textsc{gray}-channel
yielding the categories [\textsc{light}] and [\textsc{dark}]. 
Given these categories, 
the rectangular object in the scene of 
\figref{scene1-1} would be categorised as [\textsc{right upper light}].
Of course, light, dark, left, lower, etc. are labels 
that I have given to these categories. The Talking Heads 
create categories by partitioning sensory channels but do
not use these labels internally.

@@

The {\it semantic theory} required to make the 
Talking Heads experiment work is very different from 
the classical textbook approaches to the subject, which
tend to assume that categories can be defined in the 
abstract, independent of their context of use. Grounded
processes of conceptualisation and interpretation are necessarily \todo{Word order changed to fit line.}
strongly situated and context-dependent. Whether `blue 
triangle' is going to be effective in picking out the 
intended topic depends on what else is in
the scene. If all other objects in the 
context are also blue triangles, the game will fail. If there
is only one triangle, it was not really necessary to say that 
it is blue. This situatedness and context-dependence, together with 
the non-verbal hints given by the speaker, are crucial 
keys for restricting the 
possible meanings of a form 
and they therefore help the hearer disambiguate 
utterances or figure out the meaning of an unknown form. But situatedness 
and con\-text-dependence are also major
sources of difficulty, because the same categories may
refer to different things for different agents in different
contexts. Even if both agents agree that a particular form 
names a certain category, 
they may still fail in the game if the category picks
out a different referent, for example because they see 
the context from a slightly different vantage point.

>>>>

Clearly in the development of language and meaning, 
the group and the environment matter. A child 
who grows up without a caring 
community or without sufficient environmental stimuli
never develops the rich cognitive capacities 
normal adults have. From attempts to educate {\itshape wolf children}
who grow up in isolation from a human community, or 
impaired children for whom the intensity of early 
interactions are limited, we know that 
there are critical periods
where a community and a challenging environment must be
present otherwise the child's capacities for language
are damaged for the 
rest of his or her life.\footnote{See \cite{Tager:1994}.}

But how can we sufficiently 
recreate these social conditions in experiments 
with artificial systems? 
Building colonies of physical autonomous robots
roaming the world in search of stimulating 
environments and rich interactions with
other robots is not feasible today. So how 
can we ever test seriously situated and 
socially embedded approaches to cognition?

@@

Let me make a distinction between the physical aspects of 
a cognitive agent and the mental aspects. The physical aspects include the 
agent's body, the sensors and articulators, 
the physical location, the objects in this location, and the other
agents physically present in the environment. The 
mental aspects include the agent's repertoire of behaviours, 
the brain structures and processes
performing categorisations of reality, his
memory, lexicon, grammar, and so on. 
In the case of humans, these two aspects are intimately 
connected and indivisible. We cannot teleport our mental 
faculties into another body, or into another copy of 
our body, even though there have been speculations that 
we could in the future record human 
brain states,\footnote{
Such visions of the future have been put forward
by \cite{Moravec:1995}. Neither current artificial intelligence technology nor
the state of the art in brain state recording are anywhere 
near to realising these visions.}
I believe that this will still not enable teleportation
because in human brains there is no distinction between 
hardware and software. The architecture of a human brain, 
the physical connections between cells, and the 
biochemical processes in each cell determine the brain's behaviour.
There is no separation between a {\itshape brain program} and
an interpreter that reads brain programs and 
executes them. The brain is a special-purpose 
hardware device which is unique to each individual. 
To copy such a device we would have to rebuild it physically,
atom by atom, and integrate it in an exact copy of the 
same body.

@@

A blind person who receives sight after the critical 
period for acquiring visual categorisation undergoes a 
traumatic experience, see \cite{Zeki:1993}. 
We could in principle build robot bodies
through which agents can experience their world with tactile
sensing and other robot bodies which support visual sensing, or both. 
But then an agent which had only access to tactile
sensing might suddenly find himself in a body equipped with 
vision. This pathological complication will 
be avoided by using the same robotic infrastructure
in every location, even though it would be fascinating and 
technically possible to study multi-modality in its 
own right. 
We also decided to make the robots vision-based, because visual 
sensing is one of the major sources of meaning in
human natural languages.

@@

The Guessing Game is one of the common things we do with language. 
For example, I play a similar game when I sit with a friend
at the dining table and say `could you give me the salt'. 
If she guesses correctly what I mean and hands 
me the salt, the game has succeeded. If she looks at me 
with a puzzled face (maybe she does not speak English) or hands me
the salmon instead of the salt, the game has failed. In that case, 
I can gesture in the direction of the salt and say `no, no, the {\itshape salt}
please', and then she hopefully realises and gives the salt to me. Failure 
is common in natural language dialogues and may be caused by 
many factors. For example, the salmon could indeed 
be close to the salt and my pronunciation of the word `salt'
may have sounded a bit like `salmon', perhaps because there was
loud music playing in the background or perhaps
my friend does not understand English. 
A failure is often 
an opportunity to negotiate how something will be expressed
in the future. For example, the hearer may pick up a new word
or the speaker may realise that a certain word is 
not appropriate in this particular context.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap2/figs/triangle.pdf}}
\caption{\label{triangle}Any verbal interaction
involves four entities here grouped in the
{\itshape semiotic square}. The relation between utterance
and referent always needs to be established indirectly by 
passing through perception and meaning.}
\end{figure}

@@

Every physical Talking Heads set-up features a {\itshape waiting room} in 
which agents are stored, ready to be loaded into the 
robotic infrastructure or to be teleported to another
physical site. A game starts when two agents are chosen
at random from this waiting room and loaded into 
the two physical robots. The
internal architecture of each agent is connected with 
the sensori-motor apparatus of the robot so that they can 
receive the sensory data streams recorded by the camera
and control the movements of the robots. Then the general 
control system assigns randomly the role of speaker and hearer
to each of the agents and the game can begin.

@@

The data on sensory channels are values from a continuous domain
(between 0.0 and 1.0). 
To be the basis of natural language communication, these
values must be transformed into a discrete domain. This is 
precisely the task of the conceptual layer. It 
performs massive data reduction to make infinitely
rich environments manageable. One means of categorisation
is to divide up each domain of values on a particular
sensory channel into regions and 
simply assign a category to each region, thus creating a 
discrimination tree.\is{discrimination tree} For example, 
the \textsc{hpos}-channel can be cut in two halves leading to a 
distinction between [\textsc{left}] ($0.0 ≤q \textsc{hpos} < 0.5$) and [\textsc{right}] 
($0.5 ≤q \textsc{hpos} ≤q 1.0$). The triangular object 
in \figref{scene1-1} has the value \textsc{hpos}=0.35 
and would therefore be categorised as [\textsc{left}]. Similarly, the 
\textsc{vpos}-channel can be divided in two halves yielding the 
categories [\textsc{lower}] and [\textsc{upper}], and likewise the \textsc{gray}-channel
yielding the categories [\textsc{light}] and [\textsc{dark}]. 
Given these categories, 
the rectangular object in the scene of 
\figref{scene1-1} would be categorised as [\textsc{right upper light}].
Of course, light, dark, left, lower, etc. are labels 
that I have given to these categories. The Talking Heads 
create categories by partitioning sensory channels but do
not use these labels internally.

@@

The {\itshape semantic theory} required to make the 
Talking Heads experiment work is very different from 
the classical textbook approaches to the subject, which
tend to assume that categories can be defined in the 
abstract, independent of their context of use. Grounded
processes of conceptualisation and interpretation are necessarily \todo{Word order changed to fit line.}
strongly situated and context-dependent. Whether `blue 
triangle' is going to be effective in picking out the 
intended topic depends on what else is in
the scene. If all other objects in the 
context are also blue triangles, the game will fail. If there
is only one triangle, it was not really necessary to say that 
it is blue. This situatedness and context-dependence, together with 
the non-verbal hints given by the speaker, are crucial 
keys for restricting the 
possible meanings of a form 
and they therefore help the hearer disambiguate 
utterances or figure out the meaning of an unknown form. But situatedness 
and con\-text-dependence are also major
sources of difficulty, because the same categories may
refer to different things for different agents in different
contexts. Even if both agents agree that a particular form 
names a certain category, 
they may still fail in the game if the category picks
out a different referent, for example because they see 
the context from a slightly different vantage point.
